---
title: "Assignment 5A: Airline Delays, Tidying and Transforming Data"
author: "YOUR NAME HERE"
date: today
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This report tidies and analyzes a small table of arrival delays for two airlines, Alaska and AM West, across five destinations (Los Angeles, Phoenix, San Diego, San Francisco, Seattle). The table comes from *Numbersense* by Kaiser Fung (McGraw-Hill, 2013), as given in the DATA 607 assignment. We recreate it as a CSV in its original "wide" layout (blank cells included), clean and reshape it with `tidyr` and `dplyr`, and compare the airlines' delay **percentages**, first overall and then city by city. The two comparisons disagree, and we explain why.

The data file `airline_delays.csv` was typed in by hand from the source table and lives in this GitHub repository so the report is reproducible. We checked it against an independent total: 11,000 flights (3,775 Alaska + 7,225 AM West).

# Step 1: Read the data

The CSV keeps the source layout on purpose: the airline name appears only on the first row of each pair, the first two header cells are blank, and a blank row separates the airlines.

```{r read}
library(tidyverse)

url <- "https://raw.githubusercontent.com/amine-souissi0/data607-week5/main/airline_delays.csv"
raw <- read_csv(url, show_col_types = FALSE)

raw

# Tests: 5 rows x 7 columns, 3 blank airline cells, 11,000 flights in total
stopifnot(all(dim(raw) == c(5, 7)))
stopifnot(sum(is.na(raw[[1]])) == 3)
stopifnot(sum(raw[, 3:7], na.rm = TRUE) == 11000)
```

`read_csv()` names the two blank headers `...1` and `...2` and keeps the spacer row as a row of `NA` values.

# Step 2: Clean the table and fill in the missing airline names

We name the two columns, drop the spacer row, and copy each airline name down into the blank cells below it. The order matters: if we filled first, the spacer row would inherit "ALASKA" and would no longer be entirely `NA`, so it could not be dropped.

```{r clean}
clean <- raw |>
  rename(airline = 1, status = 2) |>
  filter(!if_all(everything(), is.na)) |>       # drop the all-NA spacer row first
  fill(airline, .direction = "down")            # then fill the airline name down

clean

stopifnot(nrow(clean) == 4)
stopifnot(!anyNA(clean))
stopifnot(sum(clean[, 3:7]) == 11000)
stopifnot(all(rowSums(clean[, 3:7]) == c(3274, 501, 6438, 787)))
```

# Step 3: Transform from wide to long

Each row of the long table is one observation: an airline, a status (on time or delayed) and a city, with the number of flights.

```{r long}
flights_long <- clean |>
  pivot_longer(
    cols      = -c(airline, status),
    names_to  = "city",
    values_to = "flights"
  ) |>
  mutate(status = str_replace(status, " ", "_"))

flights_long

stopifnot(nrow(flights_long) == 20)               # 2 airlines x 2 statuses x 5 cities
stopifnot(sum(flights_long$flights) == 11000)
```

# Step 4: Compute delay percentages

We compare **percentages**, not counts, because AM West operates almost twice as many flights as Alaska. For each airline and city we put `on_time` and `delayed` side by side, then divide delayed flights by all flights. The overall rate sums the counts first and then divides; it is **not** the average of the five city percentages.

```{r rates}
by_city <- flights_long |>
  pivot_wider(names_from = status, values_from = flights) |>
  mutate(total       = on_time + delayed,
         pct_delayed = delayed / total)

overall <- by_city |>
  summarise(on_time = sum(on_time),
            delayed = sum(delayed),
            .by = airline) |>
  mutate(total       = on_time + delayed,
         pct_delayed = delayed / total)

stopifnot(nrow(by_city) == 10)
stopifnot(sum(by_city$total) == 11000)
stopifnot(isTRUE(all.equal(overall$pct_delayed, c(501 / 3775, 787 / 7225))))
```

# Comparison 1: The two airlines overall

```{r overall}
overall |>
  transmute(Airline = airline, Flights = total, Delayed = delayed,
            `Delayed %` = round(100 * pct_delayed, 2)) |>
  knitr::kable()

ggplot(overall, aes(airline, pct_delayed, fill = airline)) +
  geom_col(width = 0.55) +
  geom_text(aes(label = scales::percent(pct_delayed, accuracy = 0.1)), vjust = -0.4) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.16)) +
  scale_fill_manual(values = c("ALASKA" = "#1F4E79", "AM WEST" = "#D98E04")) +
  labs(title = "Overall share of flights delayed", x = NULL, y = "Delayed flights") +
  theme_minimal() +
  theme(legend.position = "none")
```

<!-- TODO (student): rewrite this finding in your own words. -->
Across all five cities, Alaska delayed 501 of 3,775 flights (13.3%) and AM West delayed 787 of 7,225 (10.9%). By this measure AM West looks better by about 2.4 percentage points.

# Comparison 2: The two airlines city by city

```{r by-city}
city_compare <- by_city |>
  select(city, airline, pct_delayed) |>
  pivot_wider(names_from = airline, values_from = pct_delayed) |>
  mutate(gap_pp = round(100 * (ALASKA - `AM WEST`), 1),
         better = if_else(ALASKA < `AM WEST`, "ALASKA", "AM WEST"))

city_compare

stopifnot(nrow(city_compare) == 5)
stopifnot(all(city_compare$better == "ALASKA"))
stopifnot(all(abs(city_compare$gap_pp - c(-3.3, -2.7, -5.9, -11.9, -9.1)) < 1e-9))

ggplot(by_city, aes(city, pct_delayed, fill = airline)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = scales::percent(pct_delayed, accuracy = 0.1)),
            position = position_dodge(width = 0.9), vjust = -0.3, size = 3) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.33)) +
  scale_fill_manual(values = c("ALASKA" = "#1F4E79", "AM WEST" = "#D98E04")) +
  labs(title = "Share of flights delayed, by city", x = NULL, y = "Delayed flights", fill = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

<!-- TODO (student): rewrite this finding in your own words. -->
City by city, the conclusion reverses. Alaska has a lower delay rate in all five cities, by 2.7 to 11.9 percentage points. The gap is largest in San Francisco and Seattle.

# Why the two comparisons disagree (Simpson's paradox)

An airline's overall delay rate is a **weighted average** of its city rates, and the two airlines use very different weights. The table shows how each airline's flights are spread across the cities.

```{r mix}
mix <- by_city |>
  mutate(share_of_airline = total / sum(total), .by = airline)

mix |>
  select(airline, city, share_of_airline, pct_delayed) |>
  arrange(airline, city) |>
  mutate(share_of_airline = scales::percent(share_of_airline, accuracy = 0.1),
         pct_delayed      = scales::percent(pct_delayed, accuracy = 0.1)) |>
  knitr::kable()

# The weighted sum of city rates reproduces the overall rate exactly
check <- mix |>
  summarise(x = sum(share_of_airline * pct_delayed), .by = airline) |>
  pull(x)
stopifnot(isTRUE(all.equal(check, overall$pct_delayed)))
```

Phoenix is the easiest city for both airlines, and 73% of AM West's flights go there. Seattle and San Francisco are hard for both, and 57% of Alaska's flights go to Seattle. To compare the airlines fairly, we give both the **same city mix** (each city's share of all 11,000 flights) and re-weight their own city rates:

```{r adjusted}
common_mix <- by_city |>
  summarise(n = sum(total), .by = city) |>
  mutate(weight = n / sum(n))

adjusted <- by_city |>
  left_join(common_mix, by = "city") |>
  summarise(adjusted_pct = sum(weight * pct_delayed), .by = airline)

adjusted |>
  mutate(`Actual overall %`  = round(100 * overall$pct_delayed, 2),
         `Same-mix %`        = round(100 * adjusted_pct, 2)) |>
  select(Airline = airline, `Actual overall %`, `Same-mix %`) |>
  knitr::kable()

stopifnot(isTRUE(all.equal(sum(common_mix$weight), 1)))
stopifnot(all(abs(round(100 * adjusted$adjusted_pct, 2) - c(9.21, 14.48)) < 1e-9))
```

<!-- TODO (student): rewrite this explanation in your own words. -->
The overall comparison and the city-by-city comparison disagree, which is an example of Simpson's paradox. Overall, AM West appears to have fewer delays (10.9% vs 13.3%), but Alaska has a lower delay rate in each of the five cities. The reason is the mix of destinations: 73% of AM West's flights go to Phoenix, the city with the fewest delays, while 57% of Alaska's flights go to Seattle, which has many. Once both airlines are given the same city mix, Alaska's delay rate (9.2%) is well below AM West's (14.5%).

# Conclusions and Findings

<!-- TODO (student): write these in your own words. Prompts: -->

- **Which airline performs better?** TODO: state your answer and which of the two views (overall or city by city) you trust more, and why.
- **What is the lesson about overall percentages?** TODO: what should we check before trusting an overall rate?
- **What are the limits of this data?** TODO: think about only five cities, no dates or flight distances, and what "delayed" means.
- **How would you extend or verify this?** TODO: e.g. use the full flight data, add time of day or season, or test whether the gaps are statistically significant.

# AI Use

Anthropic. (2026). *Claude Sonnet 5* [Large language model]. https://claude.ai. Accessed September 24, 2026.

Claude was used as a tutor and pair-programmer: to explain tidy data and Simpson's paradox, to draft the R code for each step (which we then explained and tested against independently calculated numbers), and to draft the text of the findings, which the authors then revised. All code in this report was reviewed and can be explained by the authors.
