---
title: "Assignment 5B: Chess Elo Calculations"
author: "YOUR NAME HERE"
date: today
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This report uses the chess tournament cross-table from Project 1 (64 players, 7 rounds) to ask which players scored more, and which scored fewer, points than their ratings predicted. For every game we compute the **expected score** with the Elo formula from the pre-tournament ratings of the two players, add up each player's expected points, and subtract them from the points they actually scored. We then list the five biggest overperformers and the five biggest underperformers.

**Formula and sources.** The expected score of a player with rating $R_A$ against an opponent with rating $R_B$ is

$$E_A = \frac{1}{1 + 10^{(R_B - R_A)/400}}$$

Source: Wikipedia contributors, "Elo rating system", https://en.wikipedia.org/wiki/Elo_rating_system (accessed September 24, 2026). The assignment also points to the video "The Elo Rating System for Chess and Beyond" (February 15, 2019).

**Data.** `tournamentinfo.txt` is the Project 1 text file, stored in this GitHub repository so the report is reproducible.

# Step 1: Read the raw text and learn its layout

The file is not a table. Each player takes two lines (a player line and a detail line) followed by a separator line. We first label every line.

```{r read}
library(tidyverse)

url <- "https://raw.githubusercontent.com/amine-souissi0/data607-week5/main/tournamentinfo.txt"
raw_lines <- read_lines(url)

line_type <- case_when(
  str_detect(raw_lines, "^-+$")              ~ "separator",
  str_detect(raw_lines, "^\\s+\\d+ \\|")     ~ "player",
  str_detect(raw_lines, "^\\s+[A-Z]{2} \\|") ~ "detail",
  TRUE                                       ~ "header"
)
table(line_type)

# Tests: 196 lines; 64 players, each followed directly by a detail line
stopifnot(length(raw_lines) == 196)
stopifnot(sum(line_type == "player") == 64, sum(line_type == "detail") == 64)
stopifnot(all(which(line_type == "detail") == which(line_type == "player") + 1))
```

# Step 2: One row per player

Splitting each line on `|` puts every field in a predictable position. The rating is the digits after `R:` on the detail line.

```{r players}
players <- tibble(
  player_line = raw_lines[line_type == "player"],
  detail_line = raw_lines[line_type == "detail"]
) |>
  mutate(
    pair_num   = as.integer(str_trim(str_split_i(player_line, "\\|", 1))),
    name       = str_trim(str_split_i(player_line, "\\|", 2)),
    total_pts  = as.numeric(str_trim(str_split_i(player_line, "\\|", 3))),
    state      = str_trim(str_split_i(detail_line, "\\|", 1)),
    pre_rating = as.integer(str_match(str_split_i(detail_line, "\\|", 2), "R:\\s*(\\d+)")[, 2])
  )

players |> select(pair_num, name, state, total_pts, pre_rating) |> head()

stopifnot(nrow(players) == 64, all(players$pair_num == 1:64))
stopifnot(!anyNA(select(players, pair_num, name, state, total_pts, pre_rating)))
stopifnot(sum(players$total_pts) == 220, sum(players$pre_rating) == 88224)
stopifnot(players$name[which.min(players$pre_rating)] == "JACOB ALEXANDER LAVALLEY")
stopifnot(players$name[which.max(players$pre_rating)] == "GARY HUA")
```

Ten players have a *provisional* rating in the file (for example `R: 1220P13`, a rating based on few games). We use the number before the `P`.

# Step 3: One row per game

A result cell such as `W  39` means "won against player 39". Cells with only a letter (`H`, `B`, `U`, `X`) are rounds without an opponent: a half-point bye, a full-point bye, an unplayed round and a forfeit.

```{r games}
games <- map(4:10, \(i) str_split_i(players$player_line, "\\|", i)) |>
  set_names(paste0("R", 1:7)) |>
  as_tibble() |>
  bind_cols(pair_num = players$pair_num) |>
  pivot_longer(cols = R1:R7, names_to = "round", names_prefix = "R",
               names_transform = as.integer, values_to = "cell") |>
  mutate(result  = str_sub(str_trim(cell), 1, 1),
         opp_num = as.integer(str_extract(cell, "\\d+"))) |>
  select(pair_num, round, result, opp_num)

table(games$result)

stopifnot(nrow(games) == 448)
stopifnot(sum(is.na(games$opp_num)) == 40)
stopifnot(all(is.na(games$opp_num) == games$result %in% c("H", "B", "U", "X")))

# Mirror test: if A beat B in round r, B must show a loss to A in round r
played <- games |> filter(!is.na(opp_num))
mirror <- played |>
  rename(pair_num = opp_num, opp_num = pair_num) |>     # swap the two columns
  mutate(mirror_result = case_when(result == "W" ~ "L", result == "L" ~ "W", TRUE ~ "D")) |>
  select(pair_num, opp_num, round, mirror_result)
joined <- inner_join(played, mirror, by = c("pair_num", "opp_num", "round"))
stopifnot(nrow(joined) == nrow(played), all(joined$result == joined$mirror_result))

# Points check: 204 points from played games + 16 from byes = 220 total points
stopifnot(sum(games$result == "W") + 0.5 * sum(games$result == "D") == 204)
stopifnot(0.5 * sum(games$result == "H") + sum(games$result %in% c("B", "X")) == 16)
```

# Step 4: Elo expected score for every game

For rounds without an opponent there is nothing to predict, so we treat them as **neutral**: the expected points equal the actual points. A bye therefore adds the same amount to both totals and cannot make a player look like an over- or underperformer.

```{r elo}
elo_expected <- function(rating, opp_rating) {
  1 / (1 + 10^((opp_rating - rating) / 400))
}

# Properties of the formula
stopifnot(elo_expected(1500, 1500) == 0.5)
stopifnot(near(elo_expected(1900, 1500), 0.9091, tol = 1e-4))
stopifnot(near(elo_expected(1900, 1500) + elo_expected(1500, 1900), 1))

points <- c(W = 1, D = 0.5, L = 0, H = 0.5, B = 1, X = 1, U = 0)

games_elo <- games |>
  left_join(select(players, pair_num, rating = pre_rating), by = "pair_num") |>
  left_join(select(players, opp_num = pair_num, opp_rating = pre_rating), by = "opp_num") |>
  mutate(actual   = unname(points[result]),
         expected = coalesce(elo_expected(rating, opp_rating), actual))

stopifnot(nrow(games_elo) == 448)
stopifnot(near(sum(games_elo$actual), 220), near(sum(games_elo$expected), 220))

# Hand check: Gary Hua (pair 1) expected about 5.16 points, scored 6
gary <- games_elo |> filter(pair_num == 1)
stopifnot(near(sum(gary$expected), 5.1616, tol = 1e-4))
stopifnot(near(mean(gary$opp_rating), 1605.29, tol = 1e-2))   # Project 1 example: 1605
```

# Step 5: Expected and actual points per player

```{r summary}
player_summary <- games_elo |>
  summarise(games_played = sum(!is.na(opp_num)),
            actual       = sum(actual),
            expected     = sum(expected),
            .by = pair_num) |>
  left_join(select(players, pair_num, name, state, pre_rating), by = "pair_num") |>
  mutate(diff = actual - expected) |>
  select(pair_num, name, state, pre_rating, games_played, actual, expected, diff)

stopifnot(nrow(player_summary) == 64)
# Actual points must reproduce the official totals in the file
stopifnot(all(near(player_summary$actual,
                   players$total_pts[match(player_summary$pair_num, players$pair_num)])))
stopifnot(near(sum(player_summary$diff), 0))          # zero-sum: every game gives and takes the same
stopifnot(sum(player_summary$games_played < 7) == 23)
```

# Step 6: The five biggest over- and underperformers

```{r rank}
top5    <- player_summary |> slice_max(diff, n = 5)
bottom5 <- player_summary |> slice_min(diff, n = 5)

show <- function(d) {
  d |>
    transmute(Player = name, State = state, `Pre-rating` = pre_rating,
              Games = games_played, Actual = actual,
              Expected = round(expected, 2), Difference = round(diff, 2)) |>
    knitr::kable()
}
```

**Overperformers** (actual points above the expectation):

```{r top5}
show(top5)
```

**Underperformers** (actual points below the expectation):

```{r bottom5}
show(bottom5)
```

```{r chart, fig.height=4.5, fig.width=8}
extremes <- bind_rows(over = top5, under = bottom5, .id = "group")

ggplot(extremes, aes(reorder(name, diff), diff, fill = group)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c(over = "#1F4E79", under = "#C0504D"),
                    labels = c(over = "Overperformed", under = "Underperformed")) +
  labs(title = "Five biggest over- and underperformers vs. Elo expectation",
       x = NULL, y = "Actual minus expected points", fill = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r rank-tests}
stopifnot(setequal(top5$name, c("ADITYA BAJAJ", "ZACHARY JAMES HOUGHTON", "ANVIT RAO",
                                "JACOB ALEXANDER LAVALLEY", "STEFANO LEE")))
stopifnot(setequal(bottom5$name, c("LOREN SCHWIEBERT", "GEORGE AVERY JONES", "LARRY HODGE",
                                   "JARED GE", "RISHI SHETTY")))
stopifnot(all(abs(round(top5$diff, 2)    - c(4.05, 3.13, 3.06, 2.96, 2.71))   < 1e-9))
stopifnot(all(abs(round(bottom5$diff, 2) - c(-2.78, -2.52, -2.40, -2.01, -1.59)) < 1e-9))
stopifnot(sum(player_summary$diff > 0) == 28, sum(player_summary$diff < 0) == 36)
```

<!-- TODO (student): rewrite these observations in your own words. -->
The overperformers were mostly low-rated players (average rating about 1151) and the underperformers were mostly high-rated (about 1473). Some low ratings are unreliable: Jacob Alexander Lavalley's rating of 377 is marked provisional (based on only 3 games), and so is Zachary James Houghton's 1220. Aditya Bajaj's 1384 is not provisional and his rating rose to 1640 after the tournament, so his result reflects real strength beyond his old rating.

# Notes and limitations

- **Byes and forfeits** (40 rounds) have no opponent, so they are treated as neutral (expected points equal actual points).
- **Provisional ratings** (10 players) are noisy; large differences for those players partly reflect a bad starting rating.
- **Games played differ**: 23 players played fewer than 7 games, so they cannot build up large positive or negative differences.
- **Formula variants**: other implementations of Elo differ in small details; this report uses the standard logistic formula with a scale of 400.

# Appendix: all 64 players

```{r all}
player_summary |>
  arrange(desc(diff)) |>
  transmute(Player = name, State = state, `Pre-rating` = pre_rating, Games = games_played,
            Actual = actual, Expected = round(expected, 2), Difference = round(diff, 2)) |>
  knitr::kable()
```

# Conclusions and Findings

<!-- TODO (student): write these in your own words. Prompts: -->

- **What did the analysis show?** TODO: name the biggest over- and underperformer and explain what "expected score" means.
- **Why might low-rated players overperform?** TODO: think about provisional ratings, improving players, and regression to the mean.
- **What are the limits?** TODO: byes, few games for some players, one tournament only, one formula.
- **How would you extend or verify this?** TODO: e.g. compare with the post-tournament ratings in the file, use several tournaments, or use a different Elo implementation.

# AI Use

Anthropic. (2026). *Claude Sonnet 5* [Large language model]. https://claude.ai. Accessed September 24, 2026.

Claude was used as a tutor and pair-programmer: to explain the Elo formula and regular expressions, to draft the R code for each step (which we then explained and tested against independently calculated numbers), and to draft the observations, which the authors then revised. All code in this report was reviewed and can be explained by the authors.
